Pangenome graph layout by HOGWILD! path-guided stochastic gradient descent

### 2023-01-23 manuscript/fig_algorithm.tex
outline finished
sketching the sketches

\begin{figure*}[!htb]
	\centering
		%\hline
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
			\label{fig:metrics_explained}
		\end{subfigure}
		\begin{subfigure}[t]{0.4\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
		\label{fig:sorting_explained}
		\end{subfigure}
		%\smallskip
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
			\label{fig:pg_sgd_update_sketch}
		\end{subfigure}
		\begin{subfigure}[t]{0.4\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
			\label{fig:fig:zipf_curve}
		\end{subfigure}
	\caption{
		Explain metrics and algorithms.
		\textbf{(a)} Sketch of 1D sorting operation. \textbf{(b)} Sketch of metrics calculation. \textbf{(c)} Term updated sketch \textbf{(d)} Zipfian distribution.
	}
	\label{fig:algorithm}
\end{figure*}

>>>>

\begin{figure*}[!htb]
	\centering
		%\hline
		\begin{subfigure}[t]{0.32\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_metrics_explained.pdf}
			\label{fig:1d_before_update}
		\end{subfigure}
		\begin{subfigure}[t]{0.32\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
		\label{fig:1d_before_update_highlighted}
		\end{subfigure}
		\begin{subfigure}[t]{0.32\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/algorithm/TODO_1D_sorting_explained.pdf}
		\label{fig:1d_after_update}
		\end{subfigure}	
		%\smallskip
		\begin{subfigure}[t]{0.32\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_pg_sgd_update_sketch.png}
			\label{fig:2d_before_update}
		\end{subfigure}
		\begin{subfigure}[t]{0.32\textwidth}
			\centering
			\caption{}
			\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
			\label{fig:2d_before_update_highlighted}
		\end{subfigure}
		\begin{subfigure}[t]{0.32\textwidth}
		\centering
		\caption{}
		\includegraphics[width=\linewidth]{fig/algorithm/TODO_zipf_curve.jpg}
		\label{fig:2d_after_update}
		\end{subfigure}	
	\caption{
		PG-SGD update operation sketches. \textbf{(a)-(c)} 1D layout sketch of an update operation moving a pair of nodes at a time. Edges are not shown. \textbf{(d)-(f)} 2D layout sketch of an update operation moving a pair of nodes at a time. Edges are drawn as black links.
		\textbf{(a)} State of the graph before the 1D update. \textbf{(b)} State of the graph before the update highlighting the discrepancy of the layout distance versus the path nucleotide distance of the blue nodes in 1D.  \textbf{(c)} State of the graph after the 1D update. \textbf{(d)} State of the graph before the 2D update. \textbf{(e)} State of the graph before the update highlighting the discrepancy of the layout distance versus the path nucleotide distance of the blue nodes in 2D. \textbf{(f)} State of the graph after the 2D update.
	}
	\label{fig:sketches}
\end{figure*}

### 2023-01-23 manuscript/main.tex
sketching the sketches

\section{Background}
A pangenome1 models the entire set of genomic elements of a given population. It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves. 
\\
In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
\\
Pangenome graphs \cite{computational2016computational}. blubb.
\\
variation graph model
\\
SGD-layout. Layout in general.
\\

@@

Fig. 1: Describe how our approach works (Fig. \ref{fig:algorithm}). Explanation of 1D graph sorting and metric with Simon's handmade figures \ref{fig:sorting_explained},\ref{fig:metrics_explained}. SGD figure \ref{fig:pg_sgd_update_sketch}. Zipfian distribution figure \ref{fig:fig:zipf_curve}.

@@

\paragraph{Performance evaluation}
Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
\\
1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
\\
2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
\input{fig_performance}
\\
\\
Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
\input{tab_performance.tex}
Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
\paragraph{Latent graph structure reveals underlying biology}
Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
We want a pipeline of sortings. 2D layout of the whole HPRC.
Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
\\
We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
However, we would have to request access to the data.
\\
\\
\input{fig_latent_graph_structure}
Table 2: Metrics of the sorted graphs in Fig. 3.
\input{tab_metrics_latent_graph_structure}
\paragraph{Bonus Section}
Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
I need to test this on a new data set I got from Erik.
I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
\input{fig_tension}

>>>>

\section{Background}
A pangenome1 models the entire set of genomic elements of a given population. It can efficiently be encoded2 in the form of a variation graph, which embeds the linear sequences of the pangenome as paths in the graphs themselves. 
\\
In pangenome graphs models, DNA sequences are incorporated as nodes with edges connecting the nodes as they occur as sequences representing the graph.
\\
Pangenome graphs \cite{computational2016computational}. blubb.
\\
variation graph model
\\
SGD-layout. Layout in general.
\\ Very light background.

@@

Fig. 1: Describe how our approach works, especially a single update operation (Fig. \ref{fig:sketches}). Explanation of 1D graph updating in Figures \ref{fig:1d_before_update}-\ref{fig:1d_after_update}. Explanation of 2D graph updating in Figures \ref{fig:2d_before_update}-\ref{fig:2d_after_update}. Zipfian distribution.

@@

% something for the supplementary
\paragraph{Performance evaluation}
Fig. 2: Performance evaluation 1D + 2D: Time + RAM by number of haplotypes (Fig.~\ref{fig:haps_time_ram}). Time + RAM by number of threads (Fig.~\ref{fig:threads_time_ram}). Full HPRC graph.
\\
1D (max. threads) vs. ALIBI; we should compare time + RAM in Fig.~\ref{fig:alibi}.
But also by our sorting goodness metrics: odgi sort + the one suggested in the ALIBI paper in Table 1.
\\
2D (max. threads) + odgi draw vs. Bandage in Fig.~\ref{fig:bandage}.
\input{fig_performance}
\\
\\
Table 1: Metrics of the ALIBI and PG-SGD graphs from Fig. 2.
\input{tab_performance.tex}
Already, the PG-SGD outperforms existing graph linearization methods like the flow procedure (\url{https://doi.org/gdw58w}) or ALIBI (\url{https://doi.org/hkv3}).
\paragraph{Latent graph structure reveals underlying biology}
Fig. 3: Cool quantitative 1D sortings and 2D layouts: biological implications.
Randomly sorted. PG-SGD sorted. Ygs sorted. Reference sorted.
We want a pipeline of sortings. 2D layout of the whole HPRC.
Chr6 HPRC HLA graph? Chr8 beta-defensin gene cluster HPRC? Whole HPRC?
\\
We could also try to build a gastric cancer pangenome graph with data from \url{https://www.nature.com/articles/s41467-022-33073-7} from up to 185 samples.
However, we would have to request access to the data.
\\
\\
\input{fig_latent_graph_structure}
Table 2: Metrics of the sorted graphs in Fig. 3.
\input{tab_metrics_latent_graph_structure}
\paragraph{Bonus Section}
Fig. 4: Detect tension. Relax a graph. Detect tension afterwards.
I need to test this on a new data set I got from Erik.
I need to establish a fixed lower boundary for the tension from which on we don't relax anymore. \\
With a high quality layout, we can measure the discrepancy of the path layout position versus the expected path nucleotide position, the “tension” of a graph.
The greater the “tension”, the greater is the possibility of a biologically meaningless alignment.
This allows us to detect telomere collapsing alignment errors and hopefully (pangenome) assembly errors, subsequently correcting them.
\input{fig_tension}
